---
title: "PFAS and Cancer Incidence"
output: 
  html_document:
    toc: true
    toc_float: true
---
In this R Markdown report, we analyze and compare three methods for quantifying Per- and Polyfluoroalkyl Substances (PFAS) levels across the 52 U.S. states. Our objective is to visualize the relationship between cancer incidence and varying pollution levels, enabling a clearer understanding of the geographic distribution and potential health impacts of PFAS exposure.

## Coding
Read in the three datasets
```{r}
library(tidyverse)
library(readxl)

# US states cancer incidence dataset
fifitystates_df = read_excel("Excel_50 states.xlsx") %>%
  janitor::clean_names()
fifitystates_df

# US zipcode dataset
zipcode_df = read_excel ("US_FIPS_Codes.xls") %>%
  janitor::clean_names()
zipcode_df

# PFAS level dataset
pfas_df= read_csv ("pfas_county_073124.csv") %>%
  janitor::clean_names() %>%
  mutate(state_zipcode= substr(county, 1, 2))
pfas_df
```


Merge the three datasets
```{r}
# left join keeps everything in the dataset that comes first 
merged_df_1 = 
  left_join(pfas_df, fifitystates_df,  by = "state_zipcode")

#check the new dataset's variables‘ names
names(merged_df_1)

#check how many distinct datasets
n_distinct(merged_df_1$state)
```

Method 1: Take the median of all the variables that start with a “median" prefix, eg. median_pfbs
Group by state code
Combine all the county codes for each state and then calculate the median

```{r}
state_medians =
  merged_df_1 %>%
  group_by(state) %>%
  summarise(across(starts_with("median"), median, na.rm = TRUE))

print(state_medians)
```

Method 2 - binerization
If one county has detected/median, call this state as detected/exposed
A state can only be called non detected when all counties are non-detected

```{r}
state_detection_status =
  merged_df_1 %>%
  group_by(state) %>%
  summarise(detection_status = ifelse(any(detected == "Yes"), "Detected", "Not Detected"))

# View the result
print(state_detection_status)

table(state_detection_status$detection_status, useNA= "always" )
```

Method 3 - categorization using threshold
Create a new column for each state by calculating the number of counties that are detected
Find the cut point
```{r}
#First determine the counts
state_detection_counts =
  merged_df_1 %>%
  group_by(state) %>%
  summarise(
    yes_detected = sum(detected == "Yes"),
    no_not_detected = sum(detected == "No"),
    total_samples = n()
  )

# calculate the proportion of detection
state_detection_counts_proportion =
  state_detection_counts %>%
  mutate(
    detection_proportion = yes_detected / total_samples
  )
```

Try out different thresholds
```{r}
# 75%
state_detection_method3_seventyfive =
  state_detection_counts %>%
  mutate(
    detection_proportion = yes_detected / total_samples,
    pollution_category = case_when(
      detection_proportion > 0.75 ~ "Highly Polluted",
      detection_proportion > 0 & detection_proportion <= 0.75 ~ "Moderately Polluted",
      detection_proportion == 0 ~ "Not Polluted at All"
    )
  )

state_detection_method3_seventyfive 
table (state_detection_method3_seventyfive $pollution_category, useNA = "always")


# 60%
state_detection_method3_sixty =
  state_detection_counts %>%
  mutate(
    detection_proportion = yes_detected / total_samples,
    pollution_category = case_when(
      detection_proportion > 0.60 ~ "Highly Polluted",
      detection_proportion > 0 & detection_proportion <= 0.60 ~ "Moderately Polluted",
      detection_proportion == 0 ~ "Not Polluted at All"
    )
  )

state_detection_method3_sixty 

table (state_detection_method3_sixty$pollution_category, useNA = "always")

# 90%
state_detection_method3_ninty =
  state_detection_counts %>%
  mutate(
    detection_proportion = yes_detected / total_samples,
    pollution_category_ninty = case_when(
      detection_proportion > 0.90 ~ "Highly Polluted",
      detection_proportion > 0 & detection_proportion <= 0.90 ~ "Moderately Polluted",
      detection_proportion == 0 ~ "Not Polluted at All"
    )
  )

state_detection_method3_ninty 
table (state_detection_method3_ninty $pollution_category_ninty, useNA = "always")


# 80%
state_detection_method3_eighty =
  state_detection_counts %>%
  mutate(
    detection_proportion = yes_detected / total_samples,
    pollution_category = case_when(
      detection_proportion > 0.80 ~ "Highly Polluted",
      detection_proportion > 0 & detection_proportion <= 0.80 ~ "Moderately Polluted",
      detection_proportion == 0 ~ "Not Polluted at All"
    )
  )

state_detection_method3_eighty 
table (state_detection_method3_eighty $pollution_category, useNA = "always") 

# summary
summary(state_detection_method3_sixty$detection_proportion)


# 50% threshold
state_detection_method3_fifty =
  state_detection_counts %>%
  mutate(
    detection_proportion = yes_detected / total_samples,
    pollution_category = case_when(
      detection_proportion > 0.50 ~ "Highly Polluted",
      detection_proportion > 0 & detection_proportion <= 0.50 ~ "Moderately Polluted",
      detection_proportion == 0 ~ "Not Polluted at All"
    )
  )

state_detection_method3_fifty 
table (state_detection_method3_fifty $pollution_category, useNA = "always") 


# 30% threshold - KEEP IT AT THIS THRESHOLD
state_detection_method3_thirty =
  state_detection_counts %>%
  mutate(
    detection_proportion = yes_detected / total_samples,
    pollution_category = case_when(
      detection_proportion > 0.30 ~ "Highly Polluted",
      detection_proportion > 0 & detection_proportion <= 0.30 ~ "Moderately Polluted",
      detection_proportion == 0 ~ "Not Polluted at All"
    )
  )

# view results
state_detection_method3_thirty 
table (state_detection_method3_thirty $pollution_category, useNA = "always") 

```

We ultimately chose a 30% threshold to ensure that each category includes at least five states, meeting the recommended minimum for categorical analysis.


## Mapping

##### Map 1 PFAS Pollution Level Across U.S. States
```{r}
library(maps)
library(leaflet)
library(ggplot2)
library(dplyr)
library(sf)
```


```{r}
# Create a PFAS pollution level map
# Load U.S. state map data using sf package
us_states <- st_as_sf(maps::map("state", plot = FALSE, fill = TRUE))

# clean the "state_detection_method3_thirty" dataset
state_detection_method3_thirty <- state_detection_method3_thirty %>%
  mutate(state = case_when(
    state == "new_hampshire" ~ "new hampshire",
    state == "new_jersey" ~ "new jersey",
    state == "new_mexico" ~ "new mexico",
    state == "new_york" ~ "new york",
    state == "north_carolina" ~ "north carolina",
    state == "north_dakota" ~ "north dakota",
    state == "rhode_island" ~ "rhode island",
    state == "south_carolina" ~ "south carolina",
    state == "south_dakota" ~ "south dakota",
    state == "west_virginia" ~ "west virginia",
    TRUE ~ state  # Keep the original name if no change is needed
  ))

# Merge state map data with pollution data
state_detection_map <- us_states %>%
  left_join(state_detection_method3_thirty, by = c("ID" = "state"))
```


```{r}
ggplot(data = state_detection_map) +
  geom_sf(aes(fill = pollution_category)) +
  scale_fill_manual(values = c("Highly Polluted" = "red", "Moderately Polluted" = "yellow", "Not Polluted at All" = "green"), na.value = "grey") +
  labs(
    title = "PFAS Pollution Levels Across U.S. States",
    fill = "Pollution Category"
  ) +
  theme_minimal()
```



##### Map 2 Cancer Incidence Rates Across U.S. States
```{r}
# Create a cancer incidence gradient map
us_states <- st_as_sf(maps::map("state", plot = FALSE, fill = TRUE))

merged_df_1 <- merged_df_1 %>%
  mutate(state = case_when(
    state == "new hampshire" ~ "new hampshire",
    state == "new jersey" ~ "new jersey",
    state == "new mexico" ~ "new mexico",
    state == "new york" ~ "new york",
    state == "north carolina" ~ "north carolina",
    state == "north dakota" ~ "north dakota",
    state == "rhode island" ~ "rhode island",
    state == "south carolina" ~ "south carolina",
    state == "south dakota" ~ "south dakota",
    state == "west virginia" ~ "west virginia",
    TRUE ~ state  # Keep other states as they are
  )) %>%
  mutate(ir = as.numeric(trimws(ir)))  # Convert 'ir' to numeric


if(any(is.na(merged_df_1$ir))) {
  warning("Some values in the 'ir' column could not be converted to numeric and were set to NA.")
}

state_detection_map <- us_states %>%
  full_join(merged_df_1, by = c("ID" = "state"))


ggplot(data = state_detection_map) +
  geom_sf(aes(fill = ir), color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue", na.value = "grey", name = "Cancer Incidence Rate") +
  labs(
    title = "Cancer Incidence Rates Across U.S. States",
    fill = "Incidence Rate"
  ) +
  theme_minimal()
```



