---
title: "Plotly"
output: 
  html_document:
    toc: true
    toc_float: true
---
We're going to look at Instacart data.

```{r}
# Load datasets
library(tidyverse)
library(p8105.datasets)
library(plotly)
```

```{r}
data(instacart)

instacart = 
  instacart %>%
  janitor::clean_names() %>%
  filter(department %in% c("produce", "frozen", "dairy eggs", "pantry", "household")) %>%
  filter(order_hour_of_day >= 8 & order_hour_of_day <= 22) %>%
  select(department, reordered, order_hour_of_day, days_since_prior_order, order_dow, add_to_cart_order) 

```


## scatterplot
to examine if there’s a relationship between the time since a customer’s last order and the likelihood of reordering. 
```{r}
instacart %>%
plot_ly(x = ~days_since_prior_order,
        y = ~reordered,
        type = 'scatter',
        mode = 'markers',
        marker = list(opacity = 0.6, size = 5)) %>%
  layout(title = "Scatter Plot of Days Since Prior Order vs Reordered",
         xaxis = list(title = "Days Since Prior Order"),
         yaxis = list(title = "Reordered (1 = Yes, 0 = No)"))

```

```{r}
instacart %>%
plot_ly(x = ~order_dow,
        y = ~order_hour_of_day,
        type = 'scatter',
        mode = 'markers',
        color = ~factor(order_dow),
        marker = list(opacity = 0.6, size = 5)) %>%
  layout(title = "Scatter Plot of Order Day of Week vs Order Hour of Day",
         xaxis = list(title = "Day of the Week"),
         yaxis = list(title = "Order Hour of Day"))
```



## boxplot
To investigate how the add-to-cart order varies across different days of the week. Specifically, it can help answer questions like:
```{r}
instacart %>%
plot_ly(x = ~order_dow,
        y = ~add_to_cart_order,
        color = ~factor(order_dow),
        type = 'box',
        colors = "viridis") %>%
  layout(title = "Distribution of Add-to-Cart Order by Day of the Week",
         xaxis = list(title = "Day of the Week"),
         yaxis = list(title = "Add to Cart Order"))
```


## barplot
To investigate which departments tend to have shorter or longer gaps between orders. This could reveal if certain types of products are purchased more frequently.
```{r}
barplot <- instacart %>%
  group_by(department) %>%
  summarize(avg_days_since_prior_order = mean(days_since_prior_order, na.rm = TRUE))


barplot %>%
  plot_ly(x = ~department,
        y = ~avg_days_since_prior_order,
        type = 'bar',
        color = ~department) %>%
  layout(title = "Average Days Since Prior Order by Department",
         xaxis = list(title = "Department"),
         yaxis = list(title = "Average Days Since Prior Order"))
```

